Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-09-01 doc/invt.tex
small corrections

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality;
the optimal formula for $\alpha(t)$
might be $c/t$ with a $c$
different from $1$\cite{
morozov2007, zhou2008, morozov2009}.
%
Here we will show that the ideal $c$
is unity, but the origin of $t$ should be shifted
to accommodate the initial error.

@@

Our test system is a one-dimensional model system, in which
$i$ represents a non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$,
so is the distribution $\rho_i$ used in Eqs.
\eqref{eq:vbar_def} and \eqref{eq:error_sum}.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$.
%
At the end of the equilibration,
the bias potential would deviate from the correct value
and our aim to see how well
the bias potential is restored
at the end of the testing schedule.
%
After the equilibration,
we reset the origin of the time, $t$, to zero,
and start the testing schedule
for $T = 10^8$ steps.
%
We repeat the above process $1000$ times (unless specified otherwise)
and average the results.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/errinvt.pdf}
  }
  \caption{
    \label{fig:errinvt}
    Error $E$ versus $\lambda$
    for the modified inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda}.
    %
    The points are results from averaging over $1000$ independent runs;
    the curves are theoretical predictions.
    %
    \note{The figure was produced by \texttt{doc/fig/errinvt.gp}.
      The data were produced by the script \texttt{data/invtcscan.py},
      and saved under \texttt{data/invt}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt}
    for the Gaussian updating scheme
    with $\sigma = 10$.
    %
    The modified schedules, produced by
    limiting the sum in Eq. \eqref{eq:q_opt}
    to the first four terms,
    became similar to the inverse-time schedule,
    Eq. \eqref{eq:alpha_invt1}.
    %
    \note{This figure was produced by \texttt{doc/fig/optacmp.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h}.
      The data were saved under the directory \texttt{data/opta}.
    }%
  }
\end{center}
\end{figure}

@@

We thank Dr. J. Ma, Dr. Y. Mei, J. Wang,
J. Drake, O. Nassar, Dr. C. Lai and Dr. S. Ou
for helpful discussions.  We gratefully acknowledge the Robert A. Welch Foundation (H-0037), the National Science Foundation (CHE-1152876) and the National Institutes of Health (GM-037657) for partial support of this work.  This research used computing resources of the National Science Foundation XSEDE grid.

@@

We can rewrite the error as
a sum of the eigenmode as
%
\begin{align}
  E
  =
  \sum_{ k = 2 }^n
    \left\langle
      \Theta_k^2
    \right\rangle
  ,
\notag
%\label{eq:error_sbin_equil}
\end{align}
%
where
%
$
  \Theta_k
  =
  \frac{ 1 } { T }
  \sum_{ t = 1 } ^ T
    \theta_k( t )
  ,
$
%
with $\theta_k(t)$ defined by Eq. \eqref{eq:theta_def}.
%and $\phi_{k i}$ by Eq. \eqref{eq:eig_orthonormal_cols}.
%\begin{align}
%\theta_k( t )
%=
%\sum_{ i = 1 }^ n
%\phi_{k i} \left[
%             \frac{ h_i(t) }
%                  { p_i    }
%             - 1
%           \right],
%\end{align}
%and
%\begin{align}
%\sum_{ k = 1 }^n
%  \phi_{k i} \, \phi_{k j}
%  = p_i \, \delta_{i j}.
%\end{align}
%
\note{Because
%
\begin{align*}
  \sum_{ k = 1 }^n
    \Theta_k^2
  &=
  \frac{ 1 } { T^2 }
  \sum_{ t, t' = 1 }^T
    \sum_{ k = 1 }^n
      \theta_k( t ) \, \theta_k( t' )
  \\
  &=
  \frac{ 1 } { T^2 }
  \sum_{ t, t' = 1 }^T
    \sum_{ i, j = 1 }^n
      \xi_i(t) \, \xi_j(t')
      %\left[
      %  \frac{ h_i(t) }
      %       { p_i    }
      %  - 1
      %\right]
      %\left[
      %  \frac{ h_j(t') }
      %       { p_j     }
      %  - 1
      %\right]
      \sum_{ k = 1 }^n
        \phi_{k \, i} \, \phi_{k \, j}
  \\
  &=
  \frac{ 1 } { T^2 }
  \sum_{ t, t' = 1 }^T
    \sum_{ i, j = 1 }^n
      \xi_i(t) \, \xi_j(t') \,
      \rho_i \, \delta_{i, j}
  \\
  &=
  \sum_{ i = 1 }^n
  \frac{ 1 } { T }
  \sum_{ t = 1 }^T \xi_i(t)
  \;
  \frac{ 1 } { T }
  \sum_{ t' = 1 }^T \xi_i(t')
  \\
  &=
  \sum_{ i = 1 }^n
    \rho_i \,
      \left(
        \frac{ H_i }
             { p_i }
        -
        \sum_{r = 1}^n \rho_r
        \frac{ H_r }
             { p_r }
      \right)^2
  ,
\end{align*}
where we have used
$$
  \frac{1}{T}
  \sum_{t = 1}^T \xi_i(t)
  =
  \frac{ H_i }
       { p_i }
  -
  \sum_{r = 1}^n \rho_r
  \frac{ H_r }
       { p_r }
  .
$$
}%
The sum starts from $k = 2$, for
by Eq. \eqref{eq:eigenmode1}, % we have
$$
\theta_1(t)
=
\sum_{ i = 1 }^n
  \phi_{1i} \, \left(
    \frac{ h_i(t) } { p_i }
    -
    \sum_{r = 1}^n
    \rho_r
    \frac{ h_r(t) } { p_r }
  \right)
= 0.
$$

@@

For the triple-bin updating scheme,
%introduced in Sec. \ref{sec:nnscheme},
we can find an analytical expression
of the optimal value of $\lambda$
in the large-$n$ and asymptotic limit.
%
In this case,
if $2 \, \nu_k > 1$ for every $k$,
the asymptotic error is given by the integral
%$\sum_i \to \frac{2 \, n}{\pi} \int_0^{\pi/2} dp$, and
%
\begin{align}
  \Err_A
  =
  \frac{ 2 \, n  }
       { \pi \, T }
  \int_0^{\pi/2}
    \frac{ \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
         {   \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
  \, dp
  .
\notag
%\label{eq:error_nn_asym_int}
\end{align}

@@

\note{The table of symbols is listed in Table \ref{tab:symbols}.
  \begin{table*}\footnotesize
  \centering
  \rowcolors{1}{white}{LightGray}
  \setlength{\tabcolsep}{4pt} % column separation
  %
  \caption{\label{tab:symbols}Table of symbols.}
  \begin{tabular}{l | p{12cm} }
    Symbol          &   Description \\
    \hline
    $\mathbf{A}$    &   Transition matrix. \\
    $\alpha$        &   Updating magnitude. \\
    $b$             &   Cutoff of the kernel of homogeneous updating schemes. \\
    $C$             &   $C \equiv \int_0^{ q(T) } M(Q) \, dQ$.  \\
    $\delta$        &   Dirac's $\delta$-function. \\
    $c$             &   Constant. \\
    $\Err$          &   Error. \\
    $\ln f$         &   Updating factor in the WL algorithm.  \\
    $\phi_{ki}$     &   Eigenvectors of the updating matrix. \\
    $g$             &   Degeneracy factor, $2$ for a periodic variable, or $1$ a non-periodic one. \\
    $\Gamma_k$      &   Integral of the autocorrelation function of mode $k$. \\
    $h_i(t)$        &   Histogram.  \\
    $H_i$           &   Average histogram.  \\
    $i, j$          &   Bin indices. \\
    $\lambda_k$     &   The $k$th eigenvalue of the updating matrix. \\
    $k, l$          &   Mode indices. \\
    $K$             &   Cutoff wave number of bandpass updating schemes.  \\
    $\kappa_k(t)$   &   Autocorrelation function of mode $k$. \\
    $M(Q)$          &   Mass function, Eq. \eqref{eq:mass_func}.   \\
    $m(Q)$          &   Mass distribution,
                        $m(Q) \equiv M(Q)/\int_0^{ q(T) } M(Q') \, dQ'$.  \\
    $\mu_i$         &   Updating kernel of a homogeneous updating scheme. \\
    $n$             &   Number of bins. \\
    $\nu_k$         &   $\nu_k \equiv \lambda_k / \lambda$. \\
    $p_i$           &   Distribution. \\
    $\pi_i$         &   (Instantaneous) distribution, Eq. \eqref{eq:pi_p_phi} \\
    $q(t)$          &   $\int_0^t \alpha(t') \, dt'$.  \\
    $Q$             &   $q(T) - q(t)$.  \\
    $\theta_k(t)$   &   Noise of mode $k$, Eq. \eqref{eq:theta_def}. \\
    $\rho_i$        &   Distribution used to define the error. \\
    $\sigma_{ij}(t)$   &   Autocorrelation function of $\zeta_i(t)$. \\
    $\sigma$        &   Width of the Gaussian updating scheme. \\
    $t, \tau$       &   Time. \\
    $T$             &   Simulation length. \\
    $u_i(q')$       &   $u_i(q') \equiv e^{\lambda \, [q - q(T)]}$. \\
    $V_i(t)$        &   Original bias potential. \\
    $v_i(t)$        &   Shifted bias potential. \\
    $\mathbf w$     &   Updating matrix. \\
    $x_i(t)$        &   Average-deducted bias potential, Eq. \eqref{eq:x_def}. \\
    $\xi_i(t)$      &   The noise, Eq. \eqref{eq:xi_def}. \\
    $\chi_i$        &   $\chi_i \equiv \ln \frac{ p_i } { p^*_i } - \sum_{ j = 1 }^n p_j \, \ln \frac{ p_j } { p^*_j }$. \\
    $y_k(t)$        &   Mode of the shifted bias potential, Eq. \eqref{eq:y_def}. \\
    $Y_k$           &   Mode of the original bias potential, Eq. \eqref{eq:Y_def}. \\
    $z$             &   Reaction coordinate. \\
    $\zeta_i(t)$    &   The noise part of the instantaneous histogram, $h_i(t)$.
  \end{tabular}
  \end{table*}
}

>>>>

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.

@@

Our test system is a one-dimensional model system, in which
$i$ represents a non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$,
so is the distribution $\rho_i$ used in Eqs.
\eqref{eq:vbar_def} and \eqref{eq:error_sum}.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$.
%
Although the bias potential is correct initially,
it would deviate from the correct value
at the end of the equilibration.
%
Our aim is to see how well
the bias potential is restored
at the end of the testing schedule.
%
After the equilibration,
we reset the origin of the time, $t$, to zero,
and start the testing schedule
for $T = 10^8$ steps.
%
We repeat the above process $1000$ times (unless specified otherwise)
and average the results.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/errinvt.pdf}
  }
  \caption{
    \label{fig:errinvt}
    Error $E$ versus parameter $\lambda$
    for the modified inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda}.
    %
    The points are results from averaging over $1000$ independent runs;
    the curves are theoretical predictions.
    %
    \note{The figure was produced by \texttt{doc/fig/errinvt.gp}.
      The data were produced by the script \texttt{data/invtcscan.py},
      and saved under \texttt{data/invt}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt}
    for the Gaussian updating scheme
    with $\sigma = 10$.
    %
    The mode-limited schedules were produced by
    limiting the sum in Eq. \eqref{eq:q_opt}
    to the first four terms.
    They became similar to the inverse-time schedule,
    Eq. \eqref{eq:alpha_invt1}.
    %
    \note{This figure was produced by \texttt{doc/fig/optacmp.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h}.
      The data were saved under the directory \texttt{data/opta}.
    }%
  }
\end{center}
\end{figure}

@@

We thank Dr. J. Ma, Dr. Y. Mei, J. Wang,
J. Drake, O. Nassar, Dr. C. Lai, Dr. S. Ou and D. Stuhlsatz
for helpful discussions.  We gratefully acknowledge the Robert A. Welch Foundation (H-0037), the National Science Foundation (CHE-1152876) and the National Institutes of Health (GM-037657) for partial support of this work.  This research used computing resources of the National Science Foundation XSEDE grid.

@@

We can rewrite the error as
a sum of the eigenmode as
%
\begin{align}
  E
  =
  \sum_{ k = 2 }^n
    \left\langle
      \Theta_k^2
    \right\rangle
  ,
\notag
%\label{eq:error_sbin_equil}
\end{align}
%
where
%
$
  \Theta_k
  =
  \frac{ 1 } { T }
  \sum_{ t = 1 } ^ T
    \theta_k( t )
  ,
$
%
with $\theta_k(t)$ defined by Eq. \eqref{eq:theta_def}.
%and $\phi_{k i}$ by Eq. \eqref{eq:eig_orthonormal_cols}.
%\begin{align}
%\theta_k( t )
%=
%\sum_{ i = 1 }^ n
%\phi_{k i} \left[
%             \frac{ h_i(t) }
%                  { p_i    }
%             - 1
%           \right],
%\end{align}
%and
%\begin{align}
%\sum_{ k = 1 }^n
%  \phi_{k i} \, \phi_{k j}
%  = p_i \, \delta_{i j}.
%\end{align}
%
\note{Because
%
\begin{align*}
  \sum_{ k = 1 }^n
    \Theta_k^2
  &=
  \frac{ 1 } { T^2 }
  \sum_{ t, t' = 1 }^T
    \sum_{ k = 1 }^n
      \theta_k( t ) \, \theta_k( t' )
  \\
  &=
  \frac{ 1 } { T^2 }
  \sum_{ t, t' = 1 }^T
    \sum_{ i, j = 1 }^n
      \xi_i(t) \, \xi_j(t')
      %\left[
      %  \frac{ h_i(t) }
      %       { p_i    }
      %  - 1
      %\right]
      %\left[
      %  \frac{ h_j(t') }
      %       { p_j     }
      %  - 1
      %\right]
      \sum_{ k = 1 }^n
        \phi_{k \, i} \, \phi_{k \, j}
  \\
  &=
  \frac{ 1 } { T^2 }
  \sum_{ t, t' = 1 }^T
    \sum_{ i, j = 1 }^n
      \xi_i(t) \, \xi_j(t') \,
      \rho_i \, \delta_{i, j}
  \\
  &=
  \sum_{ i = 1 }^n
  \rho_i \,
  \frac{ 1 } { T }
  \sum_{ t = 1 }^T \xi_i(t)
  \;
  \frac{ 1 } { T }
  \sum_{ t' = 1 }^T \xi_i(t')
  \\
  &=
  \sum_{ i = 1 }^n
    \rho_i \,
      \left(
        \frac{ H_i }
             { p_i }
        -
        \sum_{r = 1}^n \rho_r
        \frac{ H_r }
             { p_r }
      \right)^2
  ,
\end{align*}
where we have used
$$
  \frac{1}{T}
  \sum_{t = 1}^T \xi_i(t)
  =
  \frac{ H_i }
       { p_i }
  -
  \sum_{r = 1}^n \rho_r
  \frac{ H_r }
       { p_r }
  .
$$
}%
The sum starts from $k = 2$, for
by Eq. \eqref{eq:eigenmode1}, % we have
$$
\theta_1(t)
=
\sum_{ i = 1 }^n
  \phi_{1i} \, \left(
    \frac{ h_i(t) } { p_i }
    -
    \sum_{r = 1}^n
    \rho_r
    \frac{ h_r(t) } { p_r }
  \right)
= 0.
$$

@@

For the triple-bin updating scheme,
%introduced in Sec. \ref{sec:nnscheme},
we can find an analytical expression
of the optimal value of $\lambda$
in the large-$n$ and asymptotic limit.
%
In this case,
if $2 \, \nu_k > 1$ for every $k$,
the asymptotic error is implicitly given by the integral
%$\sum_i \to \frac{2 \, n}{\pi} \int_0^{\pi/2} dp$, and
%
\begin{align}
  \Err_A
  =
  \frac{ 2 \, n  }
       { \pi \, T }
  \int_0^{\pi/2}
    \frac{ \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
         {   \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
  \, dp
  .
\notag
%\label{eq:error_nn_asym_int}
\end{align}

@@

\note{The table of symbols is listed in Table \ref{tab:symbols}.
  \begin{table*}\footnotesize
  \centering
  \rowcolors{1}{white}{LightGray}
  \setlength{\tabcolsep}{4pt} % column separation
  %
  \caption{\label{tab:symbols}Table of symbols.}
  \begin{tabular}{l | p{12cm} }
    Symbol          &   Description \\
    \hline
    $\mathbf{A}$    &   Transition matrix. \\
    $\alpha$        &   Updating magnitude. \\
    $b$             &   Cutoff of the kernel of homogeneous updating schemes. \\
    $C$             &   $C \equiv \int_0^{ q(T) } M(Q) \, dQ$.  \\
    $\delta$        &   Dirac's $\delta$-function. \\
    $c$             &   Constant. \\
    $\Err, \Err_R, \Err_A$          &   Error, residual error, asymptotic error. \\
    $\ln f$         &   Updating factor in the WL algorithm.  \\
    $\phi_{ki}$     &   Eigenvectors of the updating matrix. \\
    $g$             &   Degeneracy factor, $2$ for a periodic variable, or for $1$ a non-periodic one. \\
    $\Gamma_k$      &   Integral of the autocorrelation function of mode $k$. \\
    $h_i(t)$        &   Histogram.  \\
    $H_i$           &   Average histogram.  \\
    $i, j$          &   Bin indices. \\
    $\lambda_k$     &   The $k$th eigenvalue of the updating matrix. \\
    $k, l$          &   Mode indices. \\
    $K$             &   Cutoff wave number of bandpass updating schemes.  \\
    $\kappa_k(t)$   &   Autocorrelation function of mode $k$. \\
    $M(Q)$          &   Mass function, Eq. \eqref{eq:mass_func}.   \\
    $m(Q)$          &   Mass distribution,
                        $m(Q) \equiv M(Q)/\int_0^{ q(T) } M(Q') \, dQ'$.  \\
    $\mu_i$         &   Updating kernel of a homogeneous updating scheme. \\
    $n$             &   Number of bins. \\
    $\nu_k$         &   $\nu_k \equiv \lambda_k / \lambda$. \\
    $p_i$           &   Distribution. \\
    $\pi_i$         &   (Instantaneous) distribution, Eq. \eqref{eq:pi_p_phi}. \\
    $q(t)$          &   $\int_0^t \alpha(t') \, dt'$.  \\
    $Q$             &   $q(T) - q(t)$.  \\
    $\theta_k(t)$   &   Noise of mode $k$, Eq. \eqref{eq:theta_def}. \\
    $\rho_i$        &   Distribution used to define the error. \\
    $\sigma_{ij}(t)$   &   Autocorrelation function of $\zeta_i(t)$. \\
    $\sigma$        &   Width of the Gaussian updating scheme. \\
    $t, \tau$       &   Time. \\
    $T$             &   Simulation length. \\
    $u_i(q')$       &   $u_i(q') \equiv e^{\lambda \, [q - q(T)]}$. \\
    $V_i(t)$        &   Original bias potential. \\
    $v_i(t)$        &   Shifted bias potential. \\
    $\mathbf w$     &   Updating matrix. \\
    $x_i(t)$        &   Average-deducted bias potential, Eq. \eqref{eq:x_def}. \\
    $\xi_i(t)$      &   The noise, Eq. \eqref{eq:xi_def}. \\
    $\chi_i$        &   $\chi_i \equiv \ln \frac{ p_i } { p^*_i } - \sum_{ j = 1 }^n p_j \, \ln \frac{ p_j } { p^*_j }$. \\
    $y_k(t)$        &   Mode of the shifted bias potential, Eq. \eqref{eq:y_def}. \\
    $Y_k$           &   Mode of the original bias potential, Eq. \eqref{eq:Y_def}. \\
    $z$             &   Reaction coordinate. \\
    $\zeta_i(t)$    &   The noise part of the instantaneous histogram, $h_i(t)$.
  \end{tabular}
  \end{table*}
}
